---
title: "Illgraben Exploratory 2"
output: html_notebook
---


```{r}
library(arrow)
library(dplyr)      
library(ggplot2)     
library(lubridate) 
library(tidyr)  
library(patchwork)
library(corrplot)
library(sp)          # spatial objects
library(spacetime)   # space-time data classes
library(gstat)       # variograms, kriging
library(sf)
library(terra)
library(ggrepel)
library(automap)
```


### PREPROCESSING

```{r}
df_psd <- read_parquet("df_illgraben_may_to_sept.parquet")
```
```{r}
df_anom <- read_parquet("df_illgraben_anomaly_scores.parquet")

```
```{r}
coords <- read.delim("coords.txt", sep = "|", comment.char = "#", 
                     header = FALSE, stringsAsFactors = FALSE)
head(coords)
```


```{r}
names(coords)
```
```{r}
# Manually set the column names based on the header line
col_names <- c("Network", "Station", "Location", "Channel", "Latitude", "Longitude", 
               "Elevation", "Depth", "Azimuth", "Dip", "SensorDescription", "Scale", 
               "ScaleFreq", "ScaleUnits", "SampleRate", "StartTime", "EndTime")

# Apply the column names (skip the first row since it's the header)
coords <- coords[-1, ]  # Remove the header row
names(coords) <- col_names
```

```{r}
# Numeric columns proper format
coords <- coords %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude),
    Elevation = as.numeric(Elevation)
  )
```

```{r}
# Get most recent entry for each station
station_coords <- coords %>%
  arrange(Station, desc(StartTime)) %>%
  group_by(Station) %>%
  slice(1) %>%
  ungroup() %>%
  select(Station, Latitude, Longitude, Elevation)

print("Station coordinates:")
print(station_coords)
```

### Convert to MN95/LV95 Swiss Coordinate System

```{r}
# Convert to spatial object in WGS84
stations_wgs84 <- st_as_sf(station_coords, 
                          coords = c("Longitude", "Latitude"), 
                          crs = 4326)

# Transform to MN95/LV95 (EPSG:2056)
stations_mn95 <- st_transform(stations_wgs84, crs = 2056)

# Extract MN95 coordinates
stations_mn95_coords <- stations_mn95 %>%
  mutate(
    X_mn95 = st_coordinates(.)[,1],  # Easting
    Y_mn95 = st_coordinates(.)[,2],  # Northing
    Z = Elevation
  ) %>%
  st_drop_geometry() %>%
  select(Station, X_mn95, Y_mn95, Z)

print("Coordinates in MN95 system:")
print(stations_mn95_coords)
```

### MERGE NEW COORDS
```{r}
# Check station matching
cat("Stations in PSD data:", unique(df_psd$station), "\n")
cat("Stations in anomaly data:", unique(df_anom$station), "\n")
cat("Stations with coordinates:", unique(stations_mn95_coords$Station), "\n")

# Merge coordinates with data
df_psd_coords <- df_psd %>%
  left_join(stations_mn95_coords, by = c("station" = "Station"))

df_anom_coords <- df_anom %>%
  left_join(stations_mn95_coords, by = c("station" = "Station"))

# Check for successful merge
cat("Missing coordinates in PSD data:", sum(is.na(df_psd_coords$X_mn95)), "\n")
cat("Missing coordinates in anomaly data:", sum(is.na(df_anom_coords$X_mn95)), "\n")
```

### CREATE SPATIAL OBJECTS
Spatial Conversion: Converting regular data frame into a spatial data frame where each point knows its geographic location

Coordinate System: Using MN95/LV95 (EPSG:2056) - the Swiss coordinate system matches Swisstopo data

Time + Space: Each observation has both temporal (time) and spatial (X_mn95, Y_mn95, Z) information

```{r}
# Unique stations
psd_unique <- df_psd_coords %>%
distinct(station, .keep_all = TRUE) %>%
st_as_sf(coords = c("X_mn95", "Y_mn95"), crs = 2056)

# Extract coords into a data frame
psd_coords <- cbind(psd_unique, st_coordinates(psd_unique$geometry))

# st_coordinates gives columns X and Y
ggplot(psd_coords) +
  geom_point(aes(x = X, y = Y, color = Z, size = Z)) +
  ggrepel::geom_text_repel(aes(x = X, y = Y, label = station), size = 3) +
  scale_color_viridis_c(name = "Elevation (m)") +
  scale_size_continuous(name = "Elevation (m)") +
  labs(
    title = "Illgraben Station Locations",
    subtitle = "MN95/LV95 Coordinate System",
    x = "Easting (m)", y = "Northing (m)"
  ) +
  theme_minimal()
```


### VARIOGRAM
```{r}
df_anom_coords_debflow <- df_anom_coords %>% 'debris flow'=1
```

```{r}
# COMPUTE STATION-AVERAGE ANOMALY 
df_anom_coords_debflow <- df_anom_coords %>% 'debris flow'=1
df_station_avg <- df_anom_coords %>%
  group_by(station) %>%
  summarise(
    anomaly = mean(anomaly_score, na.rm = TRUE),  # average across all time points
    X_mn95 = first(X_mn95),
    Y_mn95 = first(Y_mn95),
    .groups = "drop"
  )

cat("Number of stations:", nrow(df_station_avg), "\n")
print(df_station_avg)

# CONVERT TO SPATIAL OBJECT 
time_slice <- df_station_avg
coordinates(time_slice) <- ~X_mn95 + Y_mn95
proj4string(time_slice) <- CRS("+init=epsg:2056")  # MN95/LV95

# COMPUTE VARIOGRAM
vg_model <- autofitVariogram(anomaly ~ 1, time_slice)

# EXTRACT VARIOGRAM PARAMETERS
vg_params <- list(
  nugget = vg_model$var_model$psill[1],
  sill   = sum(vg_model$var_model$psill),
  range  = vg_model$var_model$range[2],
  model  = vg_model$var_model$model[2]
)

cat("Variogram parameters (station-average):\n")
print(vg_params)

# PLOT STATION LOCATIONS WITH AVERAGE ANOMALY 
ggplot(df_station_avg) +
  geom_point(aes(x = X_mn95, y = Y_mn95, color = anomaly, size = anomaly)) +
  geom_text_repel(aes(x = X_mn95, y = Y_mn95, label = station)) +
  scale_color_viridis_c(name = "Average Anomaly") +
  scale_size_continuous(name = "Average Anomaly") +
  labs(
    title = "Station Locations with Average Anomaly",
    x = "Easting (X_mn95, m)",
    y = "Northing (Y_mn95, m)"
  ) +
  theme_minimal()

```



```{r}

# --- Function to compute temporal variogram using autofitVariogram ---
compute_autofit_temporal_variogram <- function(df_station, station_name, cutoff_hours = 24, lag_hours = 0.5) {
  
  # Filter and bin data to 5-minute intervals
  df_binned <- df_station %>%
    filter(station == station_name) %>%
    arrange(time) %>%
    mutate(time_bin = floor_date(time, "5 minutes")) %>%
    group_by(time_bin) %>%
    summarise(
      anomaly_score = mean(anomaly_score, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      # Convert to hours since start
      time_hr = as.numeric(difftime(time_bin, min(time_bin), units = "hours")),
      dummy_y = 0
    )
  
  # Convert to spatial-like object for 1D variogram
  coordinates(df_binned) <- ~time_hr + dummy_y
  proj4string(df_binned) <- CRS(as.character(NA))
  
  # Compute empirical variogram (cutoff in hours)
  vg_emp <- variogram(anomaly_score ~ 1, df_binned, cutoff = cutoff_hours, width = lag_hours)
  
  # Fit variogram automatically
  vg_model <- autofitVariogram(anomaly_score ~ 1, df_binned, cutoff = cutoff_hours, width = lag_hours)
  
  # Extract model parameters
  vg_params <- list(
    nugget = vg_model$var_model$psill[1],
    sill   = sum(vg_model$var_model$psill),
    range  = vg_model$var_model$range[2],
    model  = vg_model$var_model$model[2]
  )
  
  cat("\nTemporal variogram parameters for", station_name, ":\n")
  print(vg_params)
  
  return(list(empirical = vg_emp, fitted = vg_model))
}

# --- Compute for ILL12 and ILL16 ---
resultILL18 <- compute_autofit_temporal_variogram(df_anom_coords, "ILL18")
resultILL17 <- compute_autofit_temporal_variogram(df_anom_coords, "ILL17")
result_ILL16 <- compute_autofit_temporal_variogram(df_anom_coords, "ILL16")
result_ILL15 <- compute_autofit_temporal_variogram(df_anom_coords, "ILL15")
result_ILL14 <- compute_autofit_temporal_variogram(df_anom_coords, "ILL14")
result_ILL13 <- compute_autofit_temporal_variogram(df_anom_coords, "ILL13")
result_ILL12 <- compute_autofit_temporal_variogram(df_anom_coords, "ILL12")
result_ILL11 <- compute_autofit_temporal_variogram(df_anom_coords, "ILL11")




```

| Station | Nugget   | Sill    | Range (hours) | Model |
| ------- | -------- | ------- | ------------- | ----- |
| ILL18   | 0.000199 | 0.00279 | 4.08          | Ste   |
| ILL17   | 0.000195 | 0.00216 | 2.42          | Ste   |
| ILL16   | 0.000121 | 0.00158 | 5.07          | Ste   |
| ILL15   | 0.000337 | 0.00431 | 4.13          | Exp   |
| ILL14   | 2.44e-05 | 0.00275 | 2.49          | Ste   |
| ILL13   | 0.000131 | 0.00310 | 10.27         | Sph   |
| ILL12   | 0.000126 | 0.00451 | 6.80          | Ste   |
| ILL11   | 7.82e-05 | 0.00219 | 9.53          | Sph   |


```{r}
# Plot variograms

plot(resultILL18$empirical, model = resultILL18$fitted$var_model,
     main = "Temporal Variogram - ILL18 (5-min bins)",
     xlab = "Lag (hours)", ylab = "Semivariance")


plot(resultILL17$empirical, model = resultILL17$fitted$var_model,
     main = "Temporal Variogram - ILL17 (5-min bins)",
     xlab = "Lag (hours)", ylab = "Semivariance")

plot(result_ILL16$empirical, model = result_ILL16$fitted$var_model,
     main = "Temporal Variogram - ILL16 (5-min bins)",
     xlab = "Lag (hours)", ylab = "Semivariance")

plot(result_ILL15$empirical, model = result_ILL15$fitted$var_model,
     main = "Temporal Variogram - ILL15 (5-min bins)",
     xlab = "Lag (hours)", ylab = "Semivariance")

plot(result_ILL14$empirical, model = result_ILL14$fitted$var_model,
     main = "Temporal Variogram - ILL14 (5-min bins)",
     xlab = "Lag (hours)", ylab = "Semivariance")

plot(result_ILL13$empirical, model = result_ILL13$fitted$var_model,
     main = "Temporal Variogram - ILL13 (5-min bins)",
     xlab = "Lag (hours)", ylab = "Semivariance")

plot(result_ILL12$empirical, model = result_ILL12$fitted$var_model,
     main = "Temporal Variogram - ILL12 (5-min bins)",
     xlab = "Lag (hours)", ylab = "Semivariance")


plot(result_ILL11$empirical, model = result_ILL11$fitted$var_model,
     main = "Temporal Variogram - ILL11 (5-min bins)",
     xlab = "Lag (hours)", ylab = "Semivariance")
```

DEBRIS FLOW TEMPORAL VARIOGRAMS

```{r}
### TEMPORAL VARIOGRAMS — DEBRIS FLOW DAYS ONLY

# Filter for debris flow days
event_days <- df_anom_coords %>%
  filter(debris_flow == 1) %>%
  mutate(date = as_date(time)) %>%
  distinct(date) %>%
  pull(date)

cat("Number of debris flow event days:", length(event_days), "\n")

# --- Filter anomaly data to only include event days ---
df_anom_event <- df_anom_coords %>%
  mutate(date = as_date(time)) %>%
  filter(date %in% event_days)

# --- Compute temporal variograms per station (event days only) ---
compute_autofit_temporal_variogram <- function(df_station, station_name, cutoff_hours = 15, lag_hours = 0.5) {
  df_binned <- df_station %>%
    filter(station == station_name) %>%
    arrange(time) %>%
    mutate(time_bin = floor_date(time, "5 minutes")) %>%
    group_by(time_bin) %>%
    summarise(
      anomaly_score = mean(anomaly_score, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      time_hr = as.numeric(difftime(time_bin, min(time_bin), units = "hours")),
      dummy_y = 0
    )

  if (nrow(df_binned) < 10) {
    message("Skipping station ", station_name, ": too few observations on event days.")
    return(NULL)
  }

  coordinates(df_binned) <- ~time_hr + dummy_y
  proj4string(df_binned) <- CRS(as.character(NA))

  vg_emp <- variogram(anomaly_score ~ 1, df_binned, cutoff = cutoff_hours, width = lag_hours)
  vg_model <- autofitVariogram(anomaly_score ~ 1, df_binned, cutoff = cutoff_hours, width = lag_hours)

  vg_params <- list(
    nugget = vg_model$var_model$psill[1],
    sill   = sum(vg_model$var_model$psill),
    range  = vg_model$var_model$range[2],
    model  = vg_model$var_model$model[2]
  )

  cat("\nTemporal variogram parameters for", station_name, "(event days):\n")
  print(vg_params)

  return(list(empirical = vg_emp, fitted = vg_model))
}

# --- Apply to selected stations ---
stations_to_plot <- c("ILL18", "ILL17", "ILL16", "ILL15", "ILL14", "ILL13", "ILL12", "ILL11")

variograms_event <- lapply(stations_to_plot, function(st) {
  compute_autofit_temporal_variogram(df_anom_event, st)
})
names(variograms_event) <- stations_to_plot


# --- Plot results ---

for (st in stations_to_plot) {
  res <- variograms_event[[st]]
  if (!is.null(res)) {
    p <- plot(res$empirical, model = res$fitted$var_model,
              main = paste("Temporal Variogram (Event Days) -", st, "(5-min bins)"),
              xlab = "Lag (hours)", ylab = "Semivariance")
    print(p)  # 👈 this forces the plot to render in loops / knitr
  }
}

```

